---
title: 'Covid-19 Overview'
author: "Yves D'hondt"
date: "28/10/2020"
output:
  html_document:
    highlight: zenburn
    theme: flatly
    toc: yes
    toc_float: yes
  word_document:
    toc: yes
---

# Initialization

First, all the necessary packages and the most recent `coronavirus` dataset have to be loaded.

```{r load-libraries, warning=FALSE, message=FALSE}
# devtools::install_github("RamiKrispin/coronavirus")
library(coronavirus)
library(dplyr)
library(ggplot2)
library(tidyverse)
```

```{r load-data, warning=FALSE, message=FALSE}
update_dataset(silence = TRUE)
data("coronavirus")
```

#  Second Wave: Belgium & Switzerland

Belgium and Switzerland are among the European countries that are hit hardest by the second wave.
Let's take a look at Belgium first.

```{r plot-belgium-overview}
# filter data for Belgium
be.coronavirus <- coronavirus %>% 
  filter(
    country == "Belgium",
    date != max(date) # in case the latest data is not yet recorded
  )

# construct Belgian plot
be.plot_overview <- 
  ggplot(
    data = be.coronavirus,
    mapping = aes(x = date, y = cases)
  ) +
  geom_line(
    mapping = aes(x = date, y = cases, group = type, col = type)
  ) +
  labs(
    title = "Daily overview of Covid-19 in Belgium",
    x = "Date",
    y = "Number of cases"
  )
  NULL

# show Belgian plot
be.plot_overview
```

As can be seen, the peak in the first wave (around May) is much lower than the most recent peak.

```{r peak-difference-belgium}
# find the peak of confirmed cases during the first wave
be.first_peak <- be.coronavirus %>% 
  filter(
    date > as.Date("2020-03-01")
    & date < as.Date("2020-06-01")
    & type == "confirmed"
  ) %>% 
  summarize(
    peak = max(cases)
  )

# find the overall peak of confirmed cases
be.overall_peak <- be.coronavirus %>% 
  filter(
    type == "confirmed"
  ) %>% 
  summarize(
    peak = max(cases)
  )

# calculate the difference between the two peaks
be.peak_comp <- be.overall_peak$peak / be.first_peak$peak
```

The current peak is much higher than the peak from the first wave. With `r I(be.first_peak$peak)` cases in the first wave's peak and `r I(be.overall_peak$peak)` cases in the most recent wave's peak, the most recent peak has around `r I(round(be.peak_comp, digits=0))` times more cases than the first peak.

Let's take a look at Switzerland next.

```{r plot-switzerland-overview}
# filter data for Switzerland
ch.coronavirus <- coronavirus %>% 
  filter(
    country == "Switzerland",
    date != max(date) # in case the latest data is not yet recorded
  )

# construct Swiss plot
ch.plot_overview <- 
  ggplot(
    data = ch.coronavirus,
    mapping = aes(x = date, y = cases)
  ) +
  geom_line(
    mapping = aes(x = date, y = cases, group = type, col = type)
  ) +
  labs(
    title = "Daily overview of Covid-19 in Switzerland",
    x = "Date",
    y = "Number of cases"
  )
  NULL

# show Swiss plot
ch.plot_overview
```

Testing seems to happen sporadically (with many days where 0 cases are recorded).
To get a better view, let's take a look at just the days when we ignore the days without any cases.

```{r plot-switzerland-overview-no-zero}

# construct Swiss plot
ch.plot_overview.no_zero <- 
  ggplot(
    data = ch.coronavirus %>% filter(cases != 0),
    mapping = aes(x = date, y = cases)
  ) +
  geom_line(
    mapping = aes(x = date, y = cases, group = type, col = type)
  ) +
  labs(
    title = "Daily overview of Covid-19 in Switzerland",
    x = "Date",
    y = "Number of cases"
  )
  NULL

# show Swiss plot
ch.plot_overview.no_zero
```

As is the case for Belgium, the peak in the first wave (around May) is much lower than the most recent peak.

```{r peak-difference-switzerland}
# find the peak of confirmed cases during the first wave
ch.first_peak <- ch.coronavirus %>% 
  filter(
    date > as.Date("2020-03-01")
    & date < as.Date("2020-06-01")
    & type == "confirmed"
  ) %>% 
  summarize(
    peak = max(cases)
  )

# find the overall peak of confirmed cases
ch.overall_peak <- ch.coronavirus %>% 
  filter(
    type == "confirmed"
  ) %>% 
  summarize(
    peak = max(cases)
  )

# calculate the difference between the two peaks
ch.peak_comp <- ch.overall_peak$peak / ch.first_peak$peak
```

The current peak is much higher than the peak from the first wave. With `r I(ch.first_peak$peak)` cases in the first wave's peak and `r I(ch.overall_peak$peak)` cases in the most recent wave's peak, the most recent peak has around `r I(round(ch.peak_comp, digits=0))` times more cases than the first peak.